---
title: "Choosing a Baby Name"
author: "Amy Paschal"
date: "2/14/2018"
output: html_document
---

### Congratulations on your impending arrival!
### Let's choose some baby names!
### Of course you want to be on point so let's see what's trending.
```{r chunk1, echo=TRUE}
classProfile <- c("character", "factor", "integer")
babyColNames <-  c("Name", "Gender", "Count")
df <- read.table("yob2016.txt", header=FALSE, sep=";", colClasses=classProfile)
names(df) <- babyColNames
summary(df)
str(df)
typo <- grep("*yyy",df$Name)
```
### Oops! Found a typo: `r df[typo,c("Name")]`
### Removing . . . 
```{r chunk2, echo=TRUE}
y2016 <- df[-typo,]
y2015 <- read.table("yob2015.txt", header=FALSE, sep=",", colClasses=classProfile)
names(y2015) <- babyColNames
```
### Well, this is interesing.
### Lot's of Z names given to exactly 5 children in 2015.
```{r chunk3, echo=TRUE}
tail(y2015,10)
final <- merge(y2015, y2016, by=c("Name","Gender"), all=FALSE)
names(final)[names(final) == "Count.x"] <- "Count.2015"
names(final)[names(final) == "Count.y"] <- "Count.2016"
```
### There certainly were a lot of popular names in 2015 and 2016.
### You have `r nrow(final)` to choose from.
### Here are the 10 most popular:
```{r chunk4, echo=TRUE}
final$Total <- final$Count.2015 + final$Count.2016
final <- final[order(final$Total, decreasing = TRUE),]
head(final, 10)
```
### Congratulations!!! It's a girl.
### Here are the 10 most popular girl's names.
### I am writing them to "itsagirl.csv" for your convenience.
```{r chunk5, echo=TRUE}
print(final[final$Gender == "F",][1:10, c("Name","Total")],  row.names = FALSE)
write.table(final[final$Gender == "F",][1:10, c("Name", "Total")], "itsagirl.csv", quote=TRUE, sep=",", col.names=TRUE, row.names=FALSE)
```
### Thank you! And enjoy your beautiful baby girl.

